---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: no
    self_contained: no
  pdf_document: default
---

# Set some global options:

tell chunks to be verbose and set the working directory:
```{r setoptions, echo=TRUE}
library("knitr")
opts_chunk$set(echo = TRUE )
#opts_knit$set(base.dir = 'Assignement1/')
```

## Loading and preprocessing the data


use read.csv to read the data and use strptime to convert date coloumn into 
date format

```{r readdata}
dat      <- read.csv(file = "activity.csv")
dat$date <- strptime(dat$date, format = "%Y-%m-%d")
```

## What is mean total number of steps taken per day?

Calculate the total number of steps taken per day (ignoring NAs)
First get the number of days, then sum up the all steps and divide by number 
of days. NAs are ignored and considered zero steps.

```{r calcmean}
ndays <-  as.numeric(round (difftime( dat$date[length(dat[,2])] , dat$date[1], 
                                      units = "days") + 1))
sum(dat$steps, na.rm = T)/ndays 
```

Show the number of steps per day in a histogram 
a histogram is the  graphical 
representation of the distribution of numerical data 
[see wikipedia](https://en.wikipedia.org/wiki/Histogram)) included are Mean and Median of the distribution.

```{r plothist}
dat.aggr <-aggregate(steps ~ as.character(date), dat, sum)
hist(dat.aggr$steps, data = dat.aggr$date, breaks = 30, col = "red", 
     xlab = "steps", main = "Histogram of steps")
abline( v = mean(dat.aggr$steps), lty = 1 , col = "blue")
text(mean(dat.aggr$steps), 8, 
     labels = paste("mean = ",as.character(round(mean(dat.aggr$steps), 2))), 
     pos = 4 , col = "blue")
abline( v = median(dat.aggr$steps), lty = 1 , col = "green")
text(median(dat.aggr$steps), 8, 
     labels = paste("median = ",as.character(round(median(dat.aggr$steps), 2))), 
     pos = 2 , col = "green")
```


## What is the average daily activity pattern?

Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r plotts}
plot(dat$date, dat$steps,type = "l", xlab = "date", ylab = "steps" , 
     main = "number of steps taken per 5-minute interval")
abline( h = mean(dat$steps, na.rm = T),  lty = 1, lwd = 5 , col = "red")
text(dat$date[7500], mean(dat$steps, na.rm = T) , 
     labels = paste("average steps per 5 min"), pos = 3 , cex = 1.5, col = "red" )
```

and plot the diurnal cycle (day has 288 5 miute intervalls):

```{r plotdiurnalcylce}
dat.aggr <- aggregate(steps ~ interval, dat, mean)
plot(dat.aggr, xlab = "5 min interval of the day", lwd = 2, type = "l", 
     col = "grey",     main = "diurnal cycle of 5 minute intervals")
nmax     <- max(dat.aggr$steps)
intmaxat <- dat.aggr$interval[dat.aggr$steps == nmax]
# calcluate hour of the day (12 five min intervals per hour)
chour    <- as.character(paste (ceiling(104/12), (104%%12) * 5, sep = ":"))
abline(v=intmaxat, col = "red" )
text(intmaxat, nmax - 45, pos = 4 , cex = 1, col = "red",
    labels = paste("max number of steps \nin intervall before: \n", chour, "hh:mm") )
```


## Imputing missing values

Calculate the number of missing values in steps: 
```{r}
sum(is.na(dat$steps))
```
copy the data and set NAs to mean:
```{r}
newdat <- dat
newdat$steps[is.na(newdat$steps)] <- mean(newdat$steps, na.rm = TRUE)
```

redo the histogram plot with the new data:

```{r plothistnew}
dat.aggr <-aggregate(steps ~ as.character(date), newdat, sum)
hist(dat.aggr$steps, data = dat.aggr$date, breaks = 30, col = "red", 
     xlab = "steps", main = "Histogram of steps")
abline( v = mean(dat.aggr$steps), lty = 1 , col = "blue")
text(mean(dat.aggr$steps), 8, 
     labels = paste("mean = ",as.character(round(mean(dat.aggr$steps), 2))), 
     pos = 4 , col = "blue")
abline( v = median(dat.aggr$steps), lty = 1 , col = "green")
text(median(dat.aggr$steps), 8, 
     labels = paste("median = ",as.character(round(median(dat.aggr$steps), 2))), 
     pos = 2 , col = "green")

```


the mean is the same as before, but the median converged towards the mean.

## Are there differences in activity patterns between weekdays and weekends?

use timeDate 
```{r weekdayend}
library(timeDate)
newdat$wd <- as.factor(ifelse(isWeekday(newdat$date), "weekday","weekend"))
newdat <- subset( newdat, select = -date )
dat.aggr <-aggregate(newdat$steps,  by= list(newdat$wd,newdat$interval),
                     FUN= mean)
names(dat.aggr) <-c("wd","interval","steps")

subwday <- subset(newdat, wd == "weekday")
subwend <- subset(newdat, wd == "weekend")

par(mfrow = c(2,1), mar=c(2,2,1,1))
plt     <- aggregate(steps ~ interval, subwday, mean)
plot(plt, xlab = "interval", ylab = "number of steps", lwd = 2, type = "l", 
     col = "grey",     main = "weekdays")

plt     <- aggregate(steps ~ interval, subwend, mean)
plot(plt, xlab = "interval", ylab = "number of steps",  lwd = 2, type = "l", 
     col = "grey",     main = "weekends")


```

